\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-08-19 paper.qmd
Updated the description

In the field of single-cell studies, a common analytical task involves clustering to identify groups of cells with similar expression profiles. NLDR is commonly used to display clusters, and help to verify the results. For example, @chen2023 illustrates the use of UMAP to identify clusters in Human Peripheral Blood Mononuclear Cells (PBMC3k). @fig-umap-author is a reproduction of the published plot. It shows three well-separated clusters.

@@

To determine whether the UMAP representation with the (hyper-)parameter choice suggested by @chen2023 preserves the original data structure, we visualize the model constructed with UMAP overlaid on the \pD{} data. The @fig-umap-author shows three well-separated clusters. However, as shown in @fig-pbmc1-sc, there is no big separation between three clusters in \pD{}. Therefore, the suggested UMAP representation (@fig-umap-author) does not accurately represent the structure(s) present in PBMC3k dataset. But, when visualizing the model-in-th-data-space some unobserved structures can be seen. Some clusters have non-linear continuity patterns and high-density patches (@fig-pbmc1-sc).

@@

Model in \gD{}, on the UMAP layout, and three views of the fit in projections from $9\text{-}D$, for the PBMC3k data ($(s_1, \ s_2) = (-0.050, \ -0.041)$, $b = 870 \  (30, \ 29)$, $m = 135$, benchmark value to remove large edges is $0.099$). (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/VqqWuE0Jj6A>).
:::

In order to find a reasonable NLDR representation for the PBMC3k dataset, the absolute error for different numbers of non-empty bins using various NLDR techniques and different (hyper-)parameter settings (@fig-pbmc-abserror) were calculated. After analyzing the results, it was found that tSNE with default (hyper-)parameter setting (perplexity: $30$) achieved the lowest error when there were $136$ number of non-empty bins. Therefore, tSNE with a perplexity value set to $30$, the default parameter setting, is considered as a reasonable representation for the PBMC3k dataset.

@@

```{r}
#| fig-cap: "MSE from UMAP with diffferent (hyper-)parameter choices and tSNE with default (hyper-)parameter setting applied to PBMC3k dataset for different number of bins along the x-axis ($b_1$). Which is the most reasonable representation of the structures present in the PBMC3k data? The residual plot have a steep slope at the beginning, indicating that a smaller $b_1$ causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of $b_1$ generates a smaller error. It was observed that for every $b_1$, the lowest error occurred with tSNE using the perplexity: $30$."
#| label: fig-pbmc-abserror
#| fig-pos: H
#| out-width: 70%

@@

As shown in @fig-tsne-suggest, there are three well-separated clusters, although they are located close to each other. Additionally, non-linear structures can also be observed within the clusters (@fig-tsne-suggest). This demonstrates that tSNE accurately captures the data structure of the PBMC3k dataset, which UMAP did not achieve.

@@

We then fit the model for tSNE, and visualize the resultant model in the \pD{} data space. The model shows a quirk, as shown in @fig-pbmc2-sc. All three clusters are connected by an edge except the small and large clusters. Because the clusters are so close in \gD{}, they attempt to maintain the structure in \pD{} as well. This is evident that tSNE with perplexity $30$ provides a reasonable representation of PBMC3k data.

>>>>

In the field of single-cell studies, a common analytical task involves clustering to identify groups of cells with similar expression profiles. NLDR is commonly used to display clusters, and help to verify the results. For example, @chen2023 illustrates the use of UMAP to identify clusters in Human Peripheral Blood Mononuclear Cells (PBMC3k). @fig-umap-author is a reproduction of the published plot. It shows three well-separated clusters with big separations.

@@

To determine whether the UMAP representation with the (hyper-)parameter choice suggested by @chen2023 preserves the original data structure, we visualize the model constructed with UMAP overlaid on the $9\text{-}D$ data. The @fig-umap-author shows three well-separated clusters. However, as shown in @fig-pbmc1-sc, there is no big separation between three clusters in $9\text{-}D$. Therefore, the suggested UMAP representation (@fig-umap-author) does not accurately represent the structure(s) present in PBMC3k dataset. But, when visualizing the model-in-th-data-space some unobserved structures can be seen. Some clusters have non-linear continuity patterns and high-density patches (@fig-pbmc1-sc).

@@

Model in \gD{}, on the UMAP layout, and three views of the fit in projections from $9\text{-}D$, for the PBMC3k data ($(s_1, \ s_2) = (-0.050, \ -0.041)$, $b = 870 \  (30, \ 29)$, $m = 135$, benchmark value to remove large edges is $0.099$) (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/VqqWuE0Jj6A>).
:::

In order to find a reasonable NLDR representation for the PBMC3k dataset, the MSE for different $b_1$ using UMAP with different (hyper-)parameter settings and tSNE with default (hyper-)parameter setting (@fig-pbmc-mse) were calculated. After analyzing the results, it was found that tSNE with default (hyper-)parameter setting (perplexity: $30$) achieved the lowest error. Therefore, tSNE with a perplexity value set to $30$, the default parameter setting, is considered as a reasonable representation for the PBMC3k dataset.

@@

```{r}
#| fig-cap: "MSE from UMAP with diffferent (hyper-)parameter choices and tSNE with default (hyper-)parameter setting applied to PBMC3k dataset for different number of bins along the x-axis ($b_1$). Which is the most reasonable representation of the structures present in the PBMC3k data? The residual plot have a steep slope at the beginning, indicating that a smaller $b_1$ causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of $b_1$ generates a smaller error. It was observed that for every $b_1$, the lowest error occurred with tSNE using the perplexity: $30$."
#| label: fig-pbmc-mse
#| fig-pos: H
#| out-width: 70%

@@

As shown in @fig-tsne-suggest, there are three well-separated clusters, although the separation between the clusters are small. Additionally, non-linear structures can also be observed within the clusters (@fig-tsne-suggest). This demonstrates that tSNE accurately captures the data structure of the PBMC3k dataset, which UMAP did not achieve.

@@

<!-- We then fit the model for tSNE, and visualize the resultant model in the $9\text{-}D$ data space. The model shows a quirk, as shown in @fig-pbmc2-sc. All three clusters are connected by an edge except the small and large clusters. Because the clusters are so close in \gD{}, they attempt to maintain the structure in \pD{} as well. This is evident that tSNE with perplexity $30$ provides a reasonable representation of PBMC3k data. -->

### 2024-08-19 paper.tex
Updated the description

In the field of single-cell studies, a common analytical task involves
clustering to identify groups of cells with similar expression profiles.
NLDR is commonly used to display clusters, and help to verify the
results. For example, \citet{chen2023} illustrates the use of UMAP to
identify clusters in Human Peripheral Blood Mononuclear Cells (PBMC3k).
Figure~\ref{fig-umap-author} is a reproduction of the published plot. It
shows three well-separated clusters.

@@

To determine whether the UMAP representation with the (hyper-)parameter
choice suggested by \citet{chen2023} preserves the original data
structure, we visualize the model constructed with UMAP overlaid on the
\pD{} data. The Figure~\ref{fig-umap-author} shows three well-separated
clusters. However, as shown in Figure~\ref{fig-pbmc1-sc}, there is no
big separation between three clusters in \pD{}. Therefore, the suggested
UMAP representation (Figure~\ref{fig-umap-author}) does not accurately
represent the structure(s) present in PBMC3k dataset. But, when
visualizing the model-in-th-data-space some unobserved structures can be
seen. Some clusters have non-linear continuity patterns and high-density
patches (Figure~\ref{fig-pbmc1-sc}).

@@

\caption{\label{fig-pbmc1-sc}Model in \gD{}, on the UMAP layout, and
three views of the fit in projections from \(9\text{-}D\), for the
PBMC3k data (\((s_1, \ s_2) = (-0.050, \ -0.041)\),
\(b = 870 \  (30, \ 29)\), \(m = 135\), benchmark value to remove large
edges is \(0.099\)). (The \textbf{langevitour} software is used to view
the data with a tour, and the full video is available at
\url{https://youtu.be/VqqWuE0Jj6A}).}

@@

In order to find a reasonable NLDR representation for the PBMC3k
dataset, the absolute error for different numbers of non-empty bins
using various NLDR techniques and different (hyper-)parameter settings
(Figure~\ref{fig-pbmc-abserror}) were calculated. After analyzing the
results, it was found that tSNE with default (hyper-)parameter setting
(perplexity: \(30\)) achieved the lowest error when there were \(136\)
number of non-empty bins. Therefore, tSNE with a perplexity value set to
\(30\), the default parameter setting, is considered as a reasonable
representation for the PBMC3k dataset.

@@

\includegraphics[width=0.7\textwidth,height=\textheight]{paper_files/figure-pdf/fig-pbmc-abserror-1.pdf}

@@

\caption{\label{fig-pbmc-abserror}MSE from UMAP with diffferent
(hyper-)parameter choices and tSNE with default (hyper-)parameter
setting applied to PBMC3k dataset for different number of bins along the
x-axis (\(b_1\)). Which is the most reasonable representation of the
structures present in the PBMC3k data? The residual plot have a steep
slope at the beginning, indicating that a smaller \(b_1\) causes a
larger amount of error. Then, the slope gradually declines or level off,
indicating that a higher number of \(b_1\) generates a smaller error. It
was observed that for every \(b_1\), the lowest error occurred with tSNE
using the perplexity: \(30\).}

@@

As shown in Figure~\ref{fig-tsne-suggest}, there are three
well-separated clusters, although they are located close to each other.
Additionally, non-linear structures can also be observed within the
clusters (Figure~\ref{fig-tsne-suggest}). This demonstrates that tSNE
accurately captures the data structure of the PBMC3k dataset, which UMAP
did not achieve.

@@

We then fit the model for tSNE, and visualize the resultant model in the
\pD{} data space. The model shows a quirk, as shown in
Figure~\ref{fig-pbmc2-sc}. All three clusters are connected by an edge
except the small and large clusters. Because the clusters are so close
in \gD{}, they attempt to maintain the structure in \pD{} as well. This
is evident that tSNE with perplexity \(30\) provides a reasonable
representation of PBMC3k data.

>>>>

In the field of single-cell studies, a common analytical task involves
clustering to identify groups of cells with similar expression profiles.
NLDR is commonly used to display clusters, and help to verify the
results. For example, \citet{chen2023} illustrates the use of UMAP to
identify clusters in Human Peripheral Blood Mononuclear Cells (PBMC3k).
Figure~\ref{fig-umap-author} is a reproduction of the published plot. It
shows three well-separated clusters with big separations.

@@

To determine whether the UMAP representation with the (hyper-)parameter
choice suggested by \citet{chen2023} preserves the original data
structure, we visualize the model constructed with UMAP overlaid on the
\(9\text{-}D\) data. The Figure~\ref{fig-umap-author} shows three
well-separated clusters. However, as shown in Figure~\ref{fig-pbmc1-sc},
there is no big separation between three clusters in \(9\text{-}D\).
Therefore, the suggested UMAP representation
(Figure~\ref{fig-umap-author}) does not accurately represent the
structure(s) present in PBMC3k dataset. But, when visualizing the
model-in-th-data-space some unobserved structures can be seen. Some
clusters have non-linear continuity patterns and high-density patches
(Figure~\ref{fig-pbmc1-sc}).

@@

\caption{\label{fig-pbmc1-sc}Model in \gD{}, on the UMAP layout, and
three views of the fit in projections from \(9\text{-}D\), for the
PBMC3k data (\((s_1, \ s_2) = (-0.050, \ -0.041)\),
\(b = 870 \  (30, \ 29)\), \(m = 135\), benchmark value to remove large
edges is \(0.099\)) (The \textbf{langevitour} software is used to view
the data with a tour, and the full video is available at
\url{https://youtu.be/VqqWuE0Jj6A}).}

@@

In order to find a reasonable NLDR representation for the PBMC3k
dataset, the MSE for different \(b_1\) using UMAP with different
(hyper-)parameter settings and tSNE with default (hyper-)parameter
setting (Figure~\ref{fig-pbmc-mse}) were calculated. After analyzing the
results, it was found that tSNE with default (hyper-)parameter setting
(perplexity: \(30\)) achieved the lowest error. Therefore, tSNE with a
perplexity value set to \(30\), the default parameter setting, is
considered as a reasonable representation for the PBMC3k dataset.

@@

\includegraphics[width=0.7\textwidth,height=\textheight]{paper_files/figure-pdf/fig-pbmc-mse-1.pdf}

@@

\caption{\label{fig-pbmc-mse}MSE from UMAP with diffferent
(hyper-)parameter choices and tSNE with default (hyper-)parameter
setting applied to PBMC3k dataset for different number of bins along the
x-axis (\(b_1\)). Which is the most reasonable representation of the
structures present in the PBMC3k data? The residual plot have a steep
slope at the beginning, indicating that a smaller \(b_1\) causes a
larger amount of error. Then, the slope gradually declines or level off,
indicating that a higher number of \(b_1\) generates a smaller error. It
was observed that for every \(b_1\), the lowest error occurred with tSNE
using the perplexity: \(30\).}

@@

As shown in Figure~\ref{fig-tsne-suggest}, there are three
well-separated clusters, although the separation between the clusters
are small. Additionally, non-linear structures can also be observed
within the clusters (Figure~\ref{fig-tsne-suggest}). This demonstrates
that tSNE accurately captures the data structure of the PBMC3k dataset,
which UMAP did not achieve.

@@
